Measuring Interaction Design

### 2015-09-25 paper.tex
final?

Developing good UIs for web or mobile applications is a
complex and expensive endeavour. One reason is the combination of
devices, interaction modalities and workflows that need to be
supported.
%
Adopting Usage Centered Development (UCD) practices is effective, as is
following established design principles~\cite{constantine99}. In
particular, one of the most effective technique is early
prototyping to explore part of the five-dimensional fidelity
prototyping space~\cite{mccurdy06,buxton07}. It is particularly effective when
paired with usability investigations based on user testing or
heuristic evaluations~\cite{preece02,rubin08}.

@@

This should not be the case, however, because answers to these
questions can be automatically computed. In this way a system can
provide important insights to a designer, and support assements of
potential user flexibility, user efficiency, error proneness, ability
to recover, compactness and consistency of a design.

@@

The literature on using state-transition networks for specifying or
analysing the behavior of UIs is vast. We conducted a systematic-style
literature review, using Google Scholar and queries with combinations
of these phrases: ``user interface'', ``path analysis'', ``user
trace'', ``interaction trace'', ``navigation'', ``markov chain'',
``markov model'', ``state transition'', ``statechart'', ``metric'',
``measure''. For each query we analyzed title and abstract of the
first 50 hits and appraised their relevance based on whether the paper
discusses approaches for measuring user interfaces in the context of
usability and on the basis of a state transition model. This resulted
in 78 full-text papers that were later on re-analyzed against the same
criterion, leading to several of papers mentioned below.
%
Many papers deal with testing user interfaces, and problems related to
generating execution traces as a means to assess coverage of a test
suite~\cite{ammann2008introduction,briand2005:test:uml}. They were
excluded from the review.

@@

Lostness~\cite{smith1996lostness,otter2000lostness} is one of the few
metrics for measuring the degree to which users become lost in the
information space, and therefore considers the notion of
\emph{deviation}. Defined specifically for hypertext systems, lostness
is a user performance measure which is afuncion of the number of
visited nodes, the number of different nodes that were visited nodes,
and the number of required nodes. This measure of efficiency is
usually applied to traces of actual users, and is argued to be
suitable for hypertext systems because the predominant task is
browsing information, rather than trying to achieve specific goals. It
is suitable therefore whe some of the following three assumptions can
be relaxed: that there is a task to complete, that there is a correct
way to carry it out, that the purpose of the system is to support
users to carry out their tasks.

@@

In~\cite{hilbert2000} several approaches to analyse streams of user
events are discussed and compared. It is interesting to realize that
this is, in a sense, the inverse problem of the one we tackle in this
paper: we want to compute a subset of the possible streams of user
events given a specification of the UI, rather than trying to abstract
general properties from streams of events.

@@

Because statechart models take advantage of abstraction features and a
rich set of connecting pseudostates, they are not suitable to be
directly processed to compute metrics. For this reason, \migtool\
first \emph{flattens} the model. Flattening is a process often used
when statecharts have to be automatically
processed~\cite{briand2005:test:uml}, and it means to produce a FSA
that is behaviorally equivalent to the original statechart, with no
hierarchy between states and no concurrent regions. In most cases this
leads to an exponential number of states and transitions in the FSA,
but because the process is completely automatic and there is no need
to manually inspect the resulting FSA, this aspect is in many cases
not relevant.  \migtool\ produces an XML representation (graphML) of
the resulting FSA, the \emph{interaction graph}. It is a directed
multigraph\footnote{A directed multigraph is a graph such that there
  are 2 or more edges that have the same end points. Cycles are paths
  that include 2 or more occurrences of the same vertex. Loops are
  edges that start and end on the same vertex. ``Geodesic path'' is a
  synonymous term with ``shortest path''.}, potentially with
cycles and loops, with edges that are labeled with the name of the
corresponding action. In the following we will be using as synonymous
the terms state/vertex, and action/edge.

@@

\migtool\ is used to process a model and a scenario
and to generate execution traces of detour order = 0,
..., H. 
For each stage, up to a maximum detour order $H$, the search algorithm
creates traces with detour order $k+1$ by collecting the set $D_k$ of states
with order $0$ up to $k$, and by identifying the neighbours ($N_k$)
of $D_k$ ($N_k$ is the set of states  not included in $D_k$ that are
connected through an edge to some state in $D_k$). Edges connecting
$D_k$ with $N_k$ represent deviations that users might follow, and
geodesic paths between $N_k$ and $D_k$ constitute recovery paths that
users might follow to complete the scenario from states in
$N_k$. These deviations and recovery paths are joined and constitute
the traces of order $k+1$.  It could happen that
for some state in $N_k$ there is no path leading to any vertex in
$D_k$: in such a case the deviation leads to a dead end that prevents
the user to complete the scenario.

@@

Figure~\ref{fig:model} shows part of the model of Gmail. At some
point, a user is viewing all the conversations of the inbox (state
\texttt{viewingConversations}); available actions include moving to
the next or previuous block of conversations, refreshing the list, or
opening a specific conversation (transition
\texttt{open(conversation)}). This transition (which is assumed to
occur when the user clicks on one of the visible conversations) leads
to the state \texttt{viewingAConversation}, where the behavior of the
system is defined by a more detailed state machine. By default the
user is viewing a conversation, but by performing the \texttt{reply}
action the UI moves to a state called \texttt{replying}, where the
body of the reply can be typed, the subject can be changed, or another
addressee can be added.

@@

Inspection of the interaction graph is not particularly useful because
even for small graphs like the one obtained from our Gmail model no
particular structure is evident that was not known from the
model. Figure~\ref{fig:gom} shows a plot using a circular layout of
the 47 states: because of the large number of cycles that exist among
states, edges form an intricated web of possible action sequences. We
believe this in practice greatly reduces usefulness of typical network
analysis metrics, such as \emph{betweenness, eccentricity, page-rank,
  eigenvalue} centrality measures.

@@

To be able to obtain results that bear upon usability, we process
further the graph, by specifying scenarios and computing traces.  The
following fragment of R code specifies the bridge sets of the choosen
scenario: it comprises, in the given order, two occurrences of
\texttt{open(conversation)}, followed by \texttt{reply}, followed by
\texttt{typeBody}, etc.: \lstset{language=R}
\begin{lstlisting}
edgesWithLabel(gmail,"open(conversation)"),
edgesWithLabel(gmail,"open(conversation)"), 
edgesWithLabel(gmail,"reply"),
edgesWithLabel(gmail,"typeBody"),
edgesWithLabel(gmail,"send"),
edgesWithLabel(gmail,"compose"),
edgesWithLabel(gmail,"addRecipient"),
edgesWithLabel(gmail,"open(files)"),
edgesWithLabel(gmail,"writeBody"),
edgesWithLabel(gmail,"writeSubject"),
edgesWithLabel(gmail,"send")
\end{lstlisting}
In plain language such a scenario means opening a conversation,
opening a second one, replying to the last message of the
second conversation by typing the body of the response, sending it, and then
composing a new message by adding a recipient, an attachment, typing
the body and subject, and finally sending it.

@@

Figure~\ref{fig:bestcase} shows the length of paths in the best case,
i.e., when users would always choose the shortest route. Gmail offers
the shortest paths across the four detour orders (for order 0 the
length is 13 steps, saving more than 13\% steps compared to other
applications; for order 3 the length is 17 steps; the other
applications are remarkably similar among them).  A plausible
interpretation is therefore that Gmail not only offers many more
error-free paths, but also gives the shortest ones. Users are given
more flexibility and more efficiency.  Because also paths with order 1
or more are the shortest ones among the four applications, Gmail also
makes users more efficient in recovering from errors.  \mytodo{result}

@@

To combine these two results, we can easily compute the frequency of
paths having different length. Figure~\ref{fig:freqopt} shows, for
each application, the frequency of order 0 paths, the frequency of
paths with length less or equal to 15 (the minimum length across the
four applications), and the frequency of optimal paths (the shortest
ones). These values show that Gmail users have the lowest probability
to hit an order 0 path (because of the relative large number of order
1 paths made available by Gmail), have the lowest probability of
hitting the shortest paths (10 times smaller than the best of the
other applications), but have the highest probability to hit a
path with length 15 or less.  Thus, the flexibility and efficiency
that can be exploited with Gmail are counterbalanced by the required
knowledge and capability of chosing an optimal path. In particular,
Gmail offers many detours of order 1 which increase flexibility for
some users and might decrease effectiveness for less knowledgeable
ones.

@@

Figure~\ref{fig:ad} shows the \emph{action density} of the four
applications, \ie\ the average number of actions per state involved
in traces, and average number of \emph{unique} actions per state. The
former is an overall measure of the number of options that are made
available by a UI, the latter can be used to analyse how many
\emph{new} options the user is presented with in any state. For our
examples, the values are all in the range between 5.9 actions/state in
the case of Gmail and 7.1 for Horde, and 1.3 unique actions for
SquirrelMail and 2 for Horde. This suggests that Gmail features a more
compact design (fewer actions to do the same things), and SquirrelMail
is even more compact when it comes to the different types of actions;
thus it could be easier to learn.

\subsection{Comparing scenarios}

The same kind of analysis can be carried out to assess how suitable a
design is for a set of different scenarios. For example, a designer
might be interested in understanding what are some quantitative
differences in replying to a message as opposed to composing a new
one.

For Horde, it turns out that composing a new message entails many more
order 0 paths (154 vs 86), and a comparable number of higher order
paths. The path length in the best case is the same across the two
scenarios, but in the average case composing has a length of 7.75
steps compared to 8.5 for reply. Probability of hitting an optimal
path is 4 times higher for compose.

This means that users will have a twice as much large choice of
correct paths when composing a new message rather than ``simply''
replying to a read one. On average, when composing, users could be
9\% more efficient, and they are 4 times more likely to do the right
things. Thus, Horde is more suitable for composing new messages than
it is for replying to existing ones.
\todo[inline]{result}

@@

Execution traces can be used also to assess what is the effect of
adding a widget or feature to an existing UI.  For example, we studied
the cruise control features of cars. One of the examples is system S,
where the driver can engage the system, and once it is engaged, speed
can be increased or decreased with small or large steps. Of course the
system can also be disengaged (by pressing the brake pedal, for
example).
System A is more elaborate, as it includes also a memory function:
when it is disengaged it remembers the current speed, which can be
recalled later on.  There are two ways to re-engage it: one
by setting a new speed, and one by recalling the previous one. In
addition,  if the car drives for more than 5 minutes at a higher
speed than the set one, system A automatically disengages.

Thus, one possible design question is ``What are the effects of adding
these functionalities?'' in a typical driving scenario where a speed
is selected, then the system is disengaged, and then the same speed
needs to be set.

System S (where we assumed that re-setting the speed is done manually
by the driver with 4 actions on ``up'' and ``down'' to
approximately get the same speed) leads to 1 order 0 path requiring 7
steps; there are 8 order 1 paths with average length  9.4. The
probability of hitting the optimal path is 0.11; action density is
1.75, and unique action density is 1.25.

On the other hand, system A has 2 order 0 paths (average length 4.5),
and 7 order 1 paths (average length 6.8). 
The optimal path probability is 0.06; action density is
2.2, and unique action density is 1.2.

In both cases the reasons for detours deal with the possibility of
disengaging the system at the wrong moment.

Therefore we can conclude that:
\begin{inparaenum}
\item system A makes users more efficient;
\item with A there are two possible ways to achieve the scenario, thus
  more flexibility is given;
\item with A the probability of doing the right thing is almost half
  of system S: it might be more difficult to do the right thing
  because more possibilities are offered;
\item system A features a more compact design, with fewer action types
  to be performed at each moment, being therefore potentially easier
  to learn and remember.
\end{inparaenum}
\todo[inline]{result}

@@

UML state diagrams provide a very expressive language, well suited to
specify behavior of UIs based on descrete events. Evidence of this can
also be found in recent OMG standards, such as
IFML~\cite{ifml13}. Even though there are fundamental limits
(inability to handle undo/redo's - because that goes beyond a finite
state representation; inability to handle customizable toolbars -
because that requires models that change at runtime; inability to
handle perceptive UIs - because they are not well suited to be
described in terms of descrete states), in many practical cases they
can be be isolated and/or left aside).

Expressivity of the modeling language means that different designers
are likely to produce different models for the same UI. As a
consequence, metrics computed by \migtool\ do depend on different
modeling choices. It is worth mentioning, though, that because of the
flattening process, several differences are reduced (for example,
those dealing with using a different hierarchy of states, or with
distributing differently concurrent regions across states), and such a
sensitivity is correspondingly reduced. At the moment, however, we do
not have hard evidence to back this claim.

Differently from other model-based approaches, such as those based on
IFML, \migtool\ uses \emph{only} a model of the behavior of the
UI. Designers do not need to cope with data modeling, nor with
decisions dealing with presentation. In a sense, \migtool\ uses only the
\emph{controller} part of the Model-View-Controller paradigm that is
adopted when developing UIs. Therefore, the designer using \migtool\ to
perform analysis is free from other concerns that in the end affect
usability, and the conclusions that are derived with \migtool\ can be
combined with other results \emph{after} the analysis is
performed. This also means that no effort needs to be directed on
developing data and presentation specifications/implementations.

In terms of validity of conclusions obtained through \migtool, we can
say that because they are devoid of user behavior assumptions (such as
preferences, skills, interpretations, ergonomic constraints) they are
very general. On the other hand, predictions based on \migtool\ metrics
are also generic because they do not consider data and presentation
aspects, nor user-related factors. For example, it is unfeasible to
use \migtool\ to predict the time needed by a user to complete a
scenario. However, \migtool\ can be used to analyze the whole
interaction design and gather data to inform more specific analyses
that could be performed, for example, with CogTool~\cite{john04cogtool}.

\migtool\ is part of a suite of model-based tools for designing,
analysing and testing user interfaces developed by the company
ANONYMIZED. \migtool\ is implemented partly in Java (model processing)
and partly in R/iGraph. Development of a web-based UI of \migtool\ is
underway; it will be freely available for research purposes.

@@

>>>>

Developing good UIs for web or mobile applications is a
complex and expensive endeavor. One reason is the combination of
devices, interaction modalities and workflows that need to be
supported.
%
Adopting Usage Centered Development (UCD) practices is effective, as is
following established design principles~\cite{constantine99}. In
particular, one of the most effective technique is early
prototyping to explore part of the five-dimensional fidelity
prototyping space~\cite{mccurdy06,buxton07}. It is particularly effective when
paired with usability investigations based on user testing or
heuristic evaluations~\cite{preece02,rubin08}.

@@

This should not be the case, however, because answers to these
questions can be automatically computed. In this way a system can
provide important insights to a designer, and support assessments of
potential user flexibility, user efficiency, error proneness, ability
to recover, compactness and consistency of a design.

@@

The literature on using state-transition networks for specifying or
analyzing the behavior of UIs is vast. We conducted a systematic-style
literature review, using Google Scholar and queries with combinations
of these phrases: ``user interface'', ``path analysis'', ``user
trace'', ``interaction trace'', ``navigation'', ``markov chain'',
``markov model'', ``state transition'', ``statechart'', ``metric'',
``measure''. For each query we analyzed title and abstract of the
first 50 hits and appraised their relevance based on whether the paper
discusses approaches for measuring user interfaces in the context of
usability and on the basis of a state transition model. This resulted
in 78 full-text papers that were later on re-analyzed against the same
criterion, leading to several of papers mentioned below.
%
Many papers deal with testing user interfaces, and problems related to
generating execution traces as a means to assess coverage of a test
suite~\cite{ammann2008introduction,briand2005:test:uml}. They were
excluded from the review.

@@

Lostness~\cite{smith1996lostness,otter2000lostness} is one of the few
metrics for measuring the degree to which users become lost in the
information space, and therefore considers the notion of
\emph{deviation}. Defined specifically for hypertext systems, lostness
is a user performance measure which is a function of the number of
visited nodes, the number of different nodes that were visited nodes,
and the number of required nodes. This measure of efficiency is
usually applied to traces of actual users, and is argued to be
suitable for hypertext systems because the predominant task is
browsing information, rather than trying to achieve specific goals. It
is suitable therefore when some of the following three assumptions can
be relaxed: that there is a task to complete, that there is a correct
way to carry it out, that the purpose of the system is to support
users to carry out their tasks.

@@

In~\cite{hilbert2000} several approaches to analyze streams of user
events are discussed and compared. It is interesting to realize that
this is, in a sense, the inverse problem of the one we tackle in this
paper: we want to compute a subset of the possible streams of user
events given a specification of the UI, rather than trying to abstract
general properties from streams of events.

@@

Because statechart models take advantage of abstraction features and a
rich set of connecting pseudostates, they are not suitable to be
directly processed to compute metrics. For this reason, \migtool\
first \emph{flattens} the model. Flattening is a process often used
when statecharts have to be automatically
processed~\cite{briand2005:test:uml}, and it means to produce a FSA
that is behaviorally equivalent to the original statechart, with no
hierarchy between states and no concurrent regions. In most cases this
leads to an exponential number of states and transitions in the FSA,
but because the process is completely automatic and there is no need
to manually inspect the resulting FSA, this aspect is in many cases
not relevant.  \migtool\ produces an XML representation (graphML) of
the resulting FSA, the \emph{interaction graph}. It is a directed
multigraph\footnote{A directed multigraph is a directed graph such
  that there are 2 or more edges that have the same end points. Cycles
  are paths that include 2 or more occurrences of the same
  vertex. Loops are edges that start and end on the same
  vertex. ``Geodesic path'' is a synonymous term with ``shortest
  path''.}, potentially with cycles and loops, with edges that are
labeled with the name of the corresponding action. In the following we
will be using as synonymous the terms state/vertex, and action/edge.

@@

\migtool\ is used to process a model and a scenario
and to generate execution traces of detour order = 0,
..., H. 
For each stage, up to a maximum detour order $H$, the search algorithm
creates traces with detour order $k+1$ by collecting the set $D_k$ of states
with order $0$ up to $k$, and by identifying the neighbors ($N_k$)
of $D_k$ ($N_k$ is the set of states  not included in $D_k$ that are
connected through an edge to some state in $D_k$). Edges connecting
$D_k$ with $N_k$ represent deviations that users might follow, and
geodesic paths between $N_k$ and $D_k$ constitute recovery paths that
users might follow to complete the scenario from states in
$N_k$. These deviations and recovery paths are joined and constitute
the traces of order $k+1$.  It could happen that
for some state in $N_k$ there is no path leading to any vertex in
$D_k$: in such a case the deviation leads to a dead end that prevents
the user to complete the scenario.

@@

Figure~\ref{fig:model} shows part of the model of Gmail. At some
point, a user is viewing all the conversations of the inbox (state
\texttt{viewingConversations}); available actions include moving to
the next or previous block of conversations, refreshing the list, or
opening a specific conversation (transition
\texttt{open(conversation)}). This transition (which is assumed to
occur when the user clicks on one of the visible conversations) leads
to the state \texttt{viewingAConversation}, where the behavior of the
system is defined by a more detailed state machine. By default the
user is viewing a conversation, but by performing the \texttt{reply}
action the UI moves to a state called \texttt{replying}, where the
body of the reply can be typed, the subject can be changed, or another
addressee can be added.

@@

Inspection of the interaction graph is not particularly useful because
even for small graphs like the one obtained from our Gmail model no
particular structure is evident that was not known from the
model. Figure~\ref{fig:gom} shows a plot using a circular layout of
the 47 states: because of the large number of cycles that exist among
states, edges form an intricate web of possible action sequences. We
believe this in practice greatly reduces usefulness of typical network
analysis metrics, such as \emph{betweenness, eccentricity, page-rank,
  eigenvalue} centrality measures.

@@

To be able to obtain results that bear upon usability, we process
further the graph, by specifying scenarios and computing traces.  The
following fragment of R code specifies the bridge sets of the chosen
scenario: it comprises, in the given order, two occurrences of
\texttt{open(conversation)}, followed by \texttt{reply}, followed by
\texttt{typeBody}, etc.: \lstset{language=R}
\begin{lstlisting}
edgesWithLabel(gmail,"open(conversation)"),
edgesWithLabel(gmail,"open(conversation)"), 
edgesWithLabel(gmail,"reply"),
edgesWithLabel(gmail,"typeBody"),
edgesWithLabel(gmail,"send"),
edgesWithLabel(gmail,"compose"),
edgesWithLabel(gmail,"addRecipient"),
edgesWithLabel(gmail,"open(files)"),
edgesWithLabel(gmail,"writeBody"),
edgesWithLabel(gmail,"writeSubject"),
edgesWithLabel(gmail,"send")
\end{lstlisting}
In plain language such a scenario means opening a conversation,
opening a second one, replying to the last message of the
second conversation by typing the body of the response, sending it, and then
composing a new message by adding a recipient, an attachment, typing
the body and subject, and finally sending it.

@@

Figure~\ref{fig:bestcase} shows the length of paths in the best case,
i.e., when users would always choose the shortest route. Gmail offers
the shortest paths across the four detour orders (for order 0 the
length is 13 steps, saving more than 13\% steps compared to other
applications; for order 3 the length is 17 steps; the other
applications are remarkably similar among them).  A plausible
interpretation is therefore that Gmail not only offers many more
error-free paths, but also gives the shortest ones. Users are given
more flexibility and more efficiency.  Because also paths with order 1
or more are the shortest ones among the four applications, Gmail also
makes users more efficient in recovering from errors.

@@

To combine these two results, we can easily compute the frequency of
paths having different length. Figure~\ref{fig:freqopt} shows, for
each application, the frequency of order 0 paths, the frequency of
paths with length less or equal to 15 (the minimum length across the
four applications), and the frequency of optimal paths (the shortest
ones). These values show that Gmail users have the lowest probability
to hit an order 0 path (because of the relative large number of order
1 paths made available by Gmail), have the lowest probability of
hitting the shortest paths (10 times smaller than the best of the
other applications), but have the highest probability to hit a
path with length 15 or less.  Thus, the flexibility and efficiency
that can be exploited with Gmail are counterbalanced by the required
knowledge and capability of choosing an optimal path. In particular,
Gmail offers many detours of order 1 which increase flexibility for
some users and might decrease effectiveness for less knowledgeable
ones.

@@

Figure~\ref{fig:ad} shows the \emph{action density} of the four
applications, \ie\ the average number of actions per state involved
in traces, and average number of \emph{unique} actions per state. The
former is an overall measure of the number of options that are made
available by a UI, the latter can be used to analyze how many
\emph{new} options the user is presented with in any state. For our
examples, the values are all in the range between 5.9 actions/state in
the case of Gmail and 7.1 for Horde, and 1.3 unique actions for
SquirrelMail and 2 for Horde. This suggests that Gmail features a more
compact design (fewer actions to do the same things), and SquirrelMail
is even more compact when it comes to the different types of actions;
thus it could be easier to learn.

\subsection{Comparing scenarios}
Previous examples show how \migtool\ can be used to compare different
UIs against the same scenario.  The same kind of analysis can be
carried out to assess how suitable a design is for different
scenarios. For example, a designer might be interested in a comparison
of replying to a message as opposed to composing a new one.

Focusing now on Horde, it turns out that composing a new message
entails many more order 0 paths (154 vs 86), and a comparable number
of higher order paths. The path length in the best case is the same
across the two scenarios, but in the average case composing has a
length of 7.75 steps compared to 8.5 for reply. Probability of hitting
an optimal path is 4 times higher for compose.

This means that users will have twice as much many choices between 
correct paths when composing a new message rather than simply
replying to a read one. On average, when composing, users could be
9\% more efficient, and they are 4 times more likely to do the right
things. Thus, Horde is more suitable for composing new messages than
it is for replying to existing ones.

@@

Execution traces can be used also to assess what is the effect of
adding a widget or feature to an existing UI.  For example, we studied
some cruise control features of cars. One of the examples is system
S\footnote{There is no need to disclose the actual brand.}, where the
driver can engage the system, and once it is engaged, speed can be
increased or decreased with small or large steps. Of course the system
can also be disengaged (by pressing the brake pedal, for example).
System A is more elaborate, as it includes also a memory function:
when it is disengaged it remembers the current speed, which can be
recalled later on.  There are two ways to re-engage it: one by setting
a new speed, and one by recalling the previous one. In addition, if
the car drives for more than 5 minutes at a higher speed than the set
one, system A automatically disengages.

Thus, one possible design question is ``What are the effects of adding
these functionalities?'' in a typical driving scenario where a speed
is selected, then the system is disengaged, and later on the same
speed needs to be set.

\begin{table}
  \centering
  \begin{tabular}{|c|r|r|r|r|r|r|r|r|}\hline
    System & $N_0$ & $L_0$ & $N_1$ & $L_1$ & p & AD & UAD \\\hline
    S  & 1 & 7 & 8 & 9.4& 11\% & 1.75 & 1.25 \\
    A  & 2 & 4.5 & 7 & 6.8 & 6\% & 2.20 & 1.20 \\\hline
  \end{tabular}\vspace{1mm}
  \caption{Comparison of the two cruise control designs. $N_i$: number
    of execution traces with detour order $i$; $L_i$: average length of
    traces with detour order $i$; $p$: probability to hit the shortest execution
    trace; AD: actions/state; UAD: unique actions/state.}
  \label{tab:data}
\end{table}

When using a scenario for S where we assume that re-setting the speed
is done manually by the driver with 4 actions on ``up'' and ``down''
to approximately get the same speed), the comparison produces the
values shown in Table~\ref{tab:data}. 
System A has 2 alternative optimal paths, their length is 4.5 (36\%
shorter than in S); both systems have a comparable number of detour 1
traces, but again their length in A is  28\% shorter. The probability
of hitting the optimal trace is however twice as large in S, which
features also a more compact design. 
In both cases detours are caused by the possibility of disengaging the
system at the wrong moment.

Therefore we can conclude that:
\begin{inparaenum}[(1)]
\item system A makes users more efficient (an increase of 56\%) for
  the considered scenario;
\item with A there are two possible ways to achieve the scenario, thus
  more flexibility is given;
\item with A the probability of doing the right thing is almost half
  of system S: it might be more difficult to do the right thing
  because more possibilities are offered;
\item system A features a more compact design, with fewer action types
  to be performed at each moment.
\end{inparaenum}

@@

UML state diagrams provide a very expressive language, well suited to
specify behavior of UIs based on discrete events. Evidence of this can
also be found in recent OMG standards, such as
IFML~\cite{ifml13}. Even though there are fundamental limits
(inability to handle undo/redo's - because that goes beyond a finite
state representation; inability to handle customizable toolbars -
because that requires models that change at runtime; inability to
handle perceptive UIs - because they are not well suited to be
described in terms of discrete states), in many practical cases they
can be be isolated and/or ignored.

Expressivity of the modeling language means that different designers
are likely to produce different models for the same UI. As a
consequence, it is possible that metrics computed by \migtool\ do depend on different
modeling choices. It is worth mentioning, though, that because of the
flattening process, several differences are reduced (for example,
those dealing with using a different hierarchy of states, or with
distributing differently concurrent regions across states), and 
sensitivity is correspondingly reduced. At the moment, however, we do
not have hard evidence to back this claim.

Differently from other model-based approaches, such as those based on
IFML, \migtool\ uses \emph{only} a model of the behavior of the
UI. Designers do not need to cope with data modeling, nor with
decisions dealing with presentation. In a sense, \migtool\ uses only the
\emph{controller} part of the Model-View-Controller paradigm that is
adopted when developing UIs. Therefore, the designer using \migtool\ to
perform analysis is free from other concerns that in the end affect
usability, and the conclusions that are derived with \migtool\ can be
combined with other results \emph{after} the analysis is
performed.

In terms of validity of conclusions obtained through \migtool, we can
say that because they are devoid of user behavior assumptions (such as
preferences, skills, interpretations, ergonomic constraints) they are
very general and conservative. On the other hand, predictions based on
\migtool\ metrics are also generic because they do not consider data
and presentation aspects, nor user-related factors. For example, it is
unfeasible to use \migtool\ to predict the time needed by a user to
complete a scenario. However, as mentioned before, \migtool\ can be
used to analyze the whole interaction design and gather data to inform
more specific analyses that could be performed, for example, with
CogTool.

At the moment \migtool\ does not use weighted edges in the interaction
graph. It is easy to extend it to search paths that minimize the total
weight, and therefore in such a way to perform analyses that are
similar to the ones suggested in~\cite{thimbleby2001markov}.

Likewise it is possible to extend \migtool\ to compute also the
lostness metric~\cite{smith1996lostness}, based on states with
different detour orders. This would give yet another objective metric
to compare different designs. We opted for non doing it in light of
the fact that lostness was designed for exploratory activities, not
for goal oriented scenarios like the ones we considered in previous
examples.

\migtool\ is part of a suite of model-based tools for designing,
analyzing and testing user interfaces developed by the company
ANONYMIZED. \migtool\ is implemented partly in Java (model processing)
and partly in R/iGraph. Development of a web-based UI of \migtool\ is
underway; it will be freely available for research purposes.

@@

We presented a method that can be used to support analysis of user
interfaces even before they are built or prototyped. UML statecharts
representing the intended behavior of user interfaces are processed
and matched to interaction scenarios, producing a number of
graph-theoretic metrics defined on possible execution traces.

We showed that with this approach one can compare different designs
against the same scenario, or the same design against different
scenarios. In both situations precise, quantitative and objective
measures can be generated regarding flexibility offered to users,
their potential efficiency, and error proneness of the user interface.

The method should not be used to draw final usability conclusions, as
it is devoid of any concerns dealing with what is presented to users,
how they could perceive that, understand that, and manipulate that.
But, because the approach can be applied also when only a
specification of the user interface is available, it supports
construction and engineering of interactive systems and could help in
iteratively simplifying or improving a design.
